\chapter{Abstract}

In today's digital age, cybergrooming poses a serious threat to the safety of children and young people and requires automatic technologies with high detection performance to support child protection. This work examines the integration of LIWC-2022 features \cite{pennebaker2022liwc} into BERT representations \cite{devlin2019bert} using a cross attention based feature fusion approach. The goal is to improve detection performance while enhancing explainability by identifying relevant LIWC features that contribute to the model's decisions. Based on a dataset consisting of complete conversations from 605 identified sexual predators collected by the Perverted Justice Foundation \cite{pj} and non grooming dialogues from PAN12 \cite{inches2012pan}, a BERT baseline was created and additionally secured against domain and length leakage using synthetic data. Afterwards, the feature fusion model was evaluated using two different LIWC 2022 feature sets. The first set included all 118 LIWC features, while the second set focused on a subset of 49 psychometric LIWC features. The results show that the feature fusion increased the F1 score by 1.7\% to 98.7\% for both feature sets, with a particularly strong improvement in precision leading to fewer false positives. In addition to the feature fusion evaluation, an analysis of the differences between grooming and non grooming conversations was conducted using LIWC features, revealing strong differences in categories related to cognitive processes, social processes and temporal orientation. Furthermore, SHAP \cite{lundberg2017shap} explainability analysis was performed to identify the most relevant LIWC features contributing to the model's decisions. The SHAP analysis showed that, on average, LIWC features contributed approximately 9.66\% when using the complete LIWC feature set and about 7.41\% for the psychometric subset to the model’s predictions. Moreover, the SHAP analysis confirmed the relevance of the high level LIWC dimensions \textit{Tone}, \textit{Authenticity}, \textit{Analytic Thinking} and \textit{Clout} as well as the broader psycholinguistic domains \textit{Affect}, \textit{Cognition} and \textit{Social Processes} for distinguishing between grooming and non grooming conversations. In conclusion, this work highlights the potential of hybrid models that combine semantic understanding with psycholinguistic insights to better identify cybergrooming behavior in online communication.